Constraining the High Density Nuclear Symmetry Energy with Pions

### 2020-05-21 chapters/ch_data_analysis.tex
Routine upload

\subsection{Gating grid noise subtraction}
The gating grid essentially is a fast switch which allows the charges of adjacent wires to reach equilibrium as fast as possible. In practice the impedance of both sides was not entirely matched properly and this caused an oscillating current to bounce back and forth in an under-dampened manner. This charge bouncing back and forth caused induced signals early in the time bucket spectrum which created and extra source of noise in the data. The ground grid shielded some of the noise, but since the path to ground was located at the ends of the TPC, the gating grid noise varied in intensity across the pads.

@@

The average noise profile was constructed by averaging all the events in the gating grid runs, pad-by-pad; each pad had a slightly different noise profile since the impedance to each pad was slightly different. It is worth noting that the small electronics noise would be effectively canceled out by averaging over thousands of events. During the first stages of the software, the gating grid noise profile closest to that run would be subtracted from the raw data canceling the gating grid noise in that run.

@@

Light charged particles (p,d,t,${}^{3}$He,${}^{4}$He), beams were produced and measured in the TPC. The magnetic and slit settings of the dipoles in the BIGRIPS spectrometer was set so that the measured momentum resolution of the beam was $\frac{\delta p}{p}$ < 1\%. Two magnetic rigidity settings were studied, with an empty target. A thick Aluminum target was used to provide a slightly lower point for part of the lower rigidity setting, effectively creating three calibration points over several particle species. The production of certain particles (t,${}^{3}$He), produced too few counts to make a good measurement, and in the high momentum rigidity setting protons could not propagate down the line and there were no counts.

Since the expected momentum resolution resulting from the spectrometer was less than 1\%, the observed momentum resolution measured by the TPC is a good measurement of the combined momentum resolution of the software and TPC (intrinsic detection) system. The momentum resolution of the TPC depends on several factors such as the particle's angle, momentum, charge, track multiplicity, etc. This calibration beam represents and ideal situation where the track was parallel to the pad plane and only one particle was measured at one time. The energy loss resolution can also be directly inferred from the measurement since each energy setting represents a monochromatic source of each particle species, which has a well defined energy loss distribution. An average momentum resolution of ~2\% and the energy loss resolution of ~ 5\% was measured for particles ranging from protons to ${}^{4}$He, over the range of momenta measured in the calibration beam as summarized in Table~\ref{tb:momresolution}.

Since magnetic dipole setting of the BIGRIPS spectrometer define the energy of each particle type we can calculate the expected momenta of each particle species measured. Small corrections to the momenta were propagated using LISE++ software which can calculate the energy loss through several materials in the beam line. These corrections resulted in a small change in the momenta. The measured momenta of the calibration beam differed significantly from the expected values as seen in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}. This effect is attributed to inhomogenatities in the magnetic field which introduces electron drift velocity in the direction of $\vec{E}\times\vec{B}$ direction. The $\vec{E}\times\vec{B}$ dift velocity causes the electron trajectories to shift toward the +x-axis in the TPC coordinates causing particles of positive charge (going in the -x-axis) to have a higher measured momenta than in reality. The disagreement in measured and expected momenta is upwards of ~5\% difference in the higher momentum calibration settings. The details of the correction technique are discussed in the later Section~\ref{sec:spacecharge} are discussed in a more general correction which also includes correction for the space charge; the same correction technique was applied here in the special case of zero space charge which is the special case of only having $\vec{E}\times\vec{B}$ components .

The values under the corrected column of seen in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}, represent the data correcting for $\vec{E}\times\vec{B}$. A significant improvement is seen in the high momentum setting going from around ~5\% disagreement to within ~1\% agreement in the corrected data. For the lower momentum settings (Tables~\cref{tb:cocktail100tar,tb:cocktail100}), protons see a slight improvement of about ~1\% where as the deuterons are over corrected in both settings. The level of agreement of the all corrected values is still within the estimated momentum resolution of the TPC.

@@

Picture of cocktail before and after ExB effect
Table of LISE++ expected cocktail energies ridigity setting of dipole magnets (reference big rips line)

\subsection{Electronics calibration}
The channel number of the electronics was calibrated by measuring the response of each channel to an input pulse supplied by a pulse generator. The pulse was injected into the ground plane of the TPC. This distributed the pulse evenly across the entire pad plane over a range of input voltages. The input voltage is plotted as a function of the measured ADC channel in Fig.~\ref{fig:gaincalib} for every channel. The small variation in each channel can be seen as the wide band around each measurement point. A linear fit is performed to get the best fit line which provides a reference line which each channel is calibrated to. The right panel shows the resulting distribution of channels after calibration. This is a relative calibration technique meant to calibrate the varying gains in each channel relative to one another.

@@

As mentioned earlier, the anode wires were separated into 14 independently biased sections. The high voltage of sections 12 and 14 were reduced during the experiment due to high currents being observed on the wires. The wire section voltages were lowered once and adjusted once again. Out of all the runs used in the analysis in this thesis \ref{tb:runList}, the anode sections 12 and 14 were lowered to \SI{1085}{\volt} for runs 2272-2371 and set to \SI{1214}{\volt} for all the other runs. By lowering the voltage on these anode wires, the gas gain is lowered as compared with all the other anode wire plane sections which operate at \SI{1460}{\volt}. To account for the drop in gain, in the software we increase the gain of the pads which lie above these anode wires. To calibrate these sections we perform a relative calibration to the high gain anode sections by comparing the energy loss values of the high and low gain sections.

@@

\subsection{Extending the dynamic range of the Electronics}
Using a TPC for measurements of HIC in nuclear physics presents a different set of challenges as opposed to higher energy experiments. Typically in higher energy experiments the charge of particles is Z$e$, where Z=1. Also the particles are traveling at higher energies in which the energy losses in a detector range from the minimum of the energy loss curve to the small logarithmic climb CITE FIGURE HERE!!!. The dynamic range of electronics in such experiments can cover a wide range of particle energies in the energy loss curve. In nuclear HICs, we are interested in cluster of nucleons typically from Z=1-3 resulting from the collision, and even higher in some applications. As seen in Eq.~\ref{eq:bb}, the energy loss is proportional to $Z^2$; the energy values of the resulting particles are at much lower energies which are in the $1/\beta^2$ region. The energy range and particle types covered by the electronics are significantly limited as they quickly reach the dynamic range limitations of the electronics as the charge of a particle increases and the velocity decreases.

Several TPCs have tried to address this issue by having regions of low and high gain, either in amplification gain or in electronics gain. Here we run into the issue that high velocity particles will have little to no signal in the low gain regions. At lower velocities, particles will deposit much charge over the low gain regions but will saturate the high gain regions, whose charge values are usually lost. Only within the dynamic range will tracks have the best momentum resolution, outside the dynamic range clusters will be missing in the sections which depend on the track velocity, lowering the momentum resolution of tracking. There are ongoing efforts in the nuclear community to develop new electronics which hope to mitigate these persistent issues in TPC electronics CITE HERE, by being able to switch to a lower gain value when the maximum range is reached. Though, it is quite useful to develop a software technique which may extend the dynamic range of TPC electronics without the use of external hardware, espeically in experiments which have already been performed with older electronics technologies.

In TPCs, the effective dynamic range can be very different from the single pad dynamic range. Typically TPCs are operated inside of a magnetic field for reconstructing momentum of each particle, which requires sub-millimeter precision in the position determination of the track path. This is achieved by averaging the charge distribution over several pads which is discussed in Section CITE HERE. Therefore the effective dynamic range is related to the relative charge values of adjacent pads (between the center pad and outside pads). For example, to measure minimum ionizing particles the signal to noise ratio of the pad with the smallest charge in the distribution should be some reasonable value, say at least 6:1. From Section CITE HERE we know the central pad in the cluster holds 80\% of the total charge, where as the two adjacent pads each hold the remaining 10\%. If we require the adjacent pads to have a signal-to-noise ratio of 6:1, then the central pad would have a signal to noise ratio of 50:1. Considering this is the signal-to-noise ratio for minimum ionizing particles, and the maximum signal-to-noise ratio is 800:1, this means the effective dynamic range in the TPC is roughly 16 times that of minimum ionizing particles. The dashed lines and vertical blue bar in Fig.~ \ref{fig:intro} are separated by a factor of 20, representing the typical effective dynamic range in a TPC. This dynamic range estimate should be regarded as approximate because the energy loss fluctuates significantly about the most probable energy loss, with a long ``Landau'' like tail, as described by Bichsel \cite{bichsel}. Nevertheless, the blue dashed lines and vertical blue bar illustrate that the range of energy losses sampled in a fixed gain readout system is limited. One can change the gain and shift the energy loss range that can be sampled, but the dynamic range itself cannot be increased.

@@

The rapid increase in the stopping powers at low momentum illustrates the degree to which the effective dynamic range can be exceeded and highlights the problems encountered in studies of intermediate HICs, in which light particles with low momenta are abundantly produced along with highly charged particles. Similar problems are encountered when TPCs are used as active targets in direct reaction studies with rare isotope beams \cite{pattpc}.

Several techniques have been employed to increase the observable range of energy losses. This can be done by lowering the electronics gain of selected readout channels, or by changing the gas amplification at the readout plane in certain areas of the TPC. In the EOS TPC \cite{eos} this was done by decreasing the voltages in select anode wires in the multi-wire readout. With the prototype Active Target TPC lowering or increasing the gain was achieved by decreasing or increasing the electric fields on selected pads within a Micromegas \cite{pattpc}. The results of changing the gas-gain, or the electronics gain, are rather similar in that reducing the gain to sample a range of higher energy loss makes the TPC effectively blind to minimum ionizing particles in  the regions of lower gain.

%DOUBLY DEFINED? IS THIS SUPPOSED TO BE HERE
%\begin{figure}[ht!]
%\includegraphics[scale=.5]{top_view_helix_ext.pdf}
%\caption{Cartoon of a top down view of a fit to a track passing through several pads. The bolded pads and the charges $q_i$ represent the hits belonging to that pad and the clusters of the track representing the average position of the track. The three clusters at the bottom are clustered in the $x$-direction for the upper three are clustered in the $z$-direction. The estimated position of the avalanche is given by the track fit, and the position from the center to each pad to the $\bar{x}$ position is given as $\lambda_i$.}
%\label{fig:topview}
%\end{figure}

We define the clustering direction depending on the angle of the track, at the point of each cluster, with respects to the $x$ axis, defined as $\theta$. For example, the crossing angle is defined as $90^{\circ}$ for a track going along the $z$ axis, and $0^{\circ}$ for a track going along the $x$ axis. In the case that the crossing angle is $45^{\circ} < \theta \leq 90^{\circ} $ the clustering direction is along the $x$ axis. For $0^{\circ} < \theta \leq 45^{\circ}$ it is along the $z$ axis.

 The position along the clustering direction is calculated by weighting the individual hit's positions by their charges $q_i$ and getting the mean value. The other direction is set to the center of the pad. For example, if we are clustering along the $x$ axis for a cluster, the $z$-position is set to the center of the pad in the $z$-direction and vice versa.

@@

The correlations we introduced by only clustering along one direction do not play a significant role in the particle identification, but cause deviations from the expected Gatti distribution. Also, analytic PRFs only exist for classical multi-wire TPCs. For these reasons it is useful to experimentally measure the PRF and fit it with an empirical function, typically a Gaussian, to describe its behavior.

As in Fig.~\ref{fig:topview}, we postulate that the PRF is a function of the total charge deposited in a cluster $Q = \sum_i q_i$, and the difference in position of the center of the $i^{th}$ pad, $x_i$, to the mean position $\bar{x} = \sum_i x_i q_i/Q$, defined as $\lambda_i = x_i-\bar{x}$. The PRF is simply defined as the charge fraction of each pad as a function of $\lambda$, as shown in Equation \ref{eq:prf}.

@@

\begin{figure}[ht!]s
\begin{overpic}[width=\linewidth]{fig5.pdf}
\put(61,55){\contour{white}{ PRF${}_{\mathrm{Gaus}}(\lambda)$ eq. \ref{eq:gaus}  }}
\put(61,49){\contour{white}{ PRF${}_{\mathrm{Gatti}}(\lambda)$ eq. \ref{eq:gatti} }}
\end{overpic}
\caption{Experimental pad response function of many events for a crossing angle of $85^{\circ} < \theta \leq 90^{\circ}$.  }
\label{fig:expprf}
\end{figure}

The shape of the PRF depends on the crossing angle of the track \cite{gatti}. Plotted in Fig.~\ref{fig:prfpimData} is the PRF of $\pi^-$ tracks vs. the crossing angle $\theta$. The PRF gets wider starting from $90^{\circ}$  and going to $45^{\circ}$; if we did not switch clustering directions the PRF would become wider until it was a uniform distribution and there was no position resolution. Since we switch the clustering direction from $x$ to the $z$ direction at $45^{\circ}$, the opposite trend is seen where the PRF becomes narrower as the position resolution gets better going from $45^{\circ}$ to $0^{\circ}$.

@@

Fits were performed to the experimental data with  $5^{\circ}$ width bins from $0^{\circ} < \theta \leq 90^{\circ}$. The two parameters of the Gaussian fits are plotted versus $\theta$ in Fig.~\ref{fig:normsigma}; a $4^{th}$ order polynomial fit between these points allowed for interpolating between $\theta$.

\subsection{Method of Desaturation}
We will use the term ``desaturation'' for our process of correcting the charge values of the saturated pads. Figure \ref{fig:satpad} shows a typical situation of saturated signals. When an avalanche causes a large induced signal, the pads directly underneath collect the largest charge becoming saturated, denoted as $q_{2'}$ and $q_{3'}$. Pads further away experience smaller, non saturating charges, denoted as $q_{1}$ and $q_{4}$. Though we do not know the saturated charge values, the distribution of all charges must follow the PRF which we have experimentally measured. From the clusters crossing angle, we can get the corresponding parameters for the PRF as described above and in Fig.~\ref{fig:normsigma}.

We assume the distance of each pad to the track, $\lambda_i$, is fixed, defining the fraction of charge each pad receives as given by the $PRF(\lambda_i)$ function.

\begin{equation}\label{eq:chi}
\chi^2 = \sum_i \frac{(q_i^{\mathrm{obs}} - q_i^{\mathrm{expect}})^2}{q_i^{\mathrm{expect}}}
\end{equation}

To determine the best estimate for the charge values of each saturated pad, a chi squared function is minimized, given in  Equation \ref{eq:chi}, where $q_i^{\mathrm{obs}}$ are the observed, non-saturated charges $q_{1}$ and $q_{4}$, and $q_i^{\mathrm{expect}}$ are the charges we expect to observe, calculated as $q_i^{\mathrm{expected}} = Q\cdot PRF(\lambda_i)$. The charges $q_{2'}$ and $q_{3'}$, make up the unknown variable and are allowed to vary in the $\chi^2$ minimization, where they are added to make up the total expected charge $Q$.

@@

Shown in Fig.~\ref{fig:cocktail} is a typical cocktail event, where one particle enters the TPC volume at a time and parallel to the pad plane, representing an ideal case for momentum and $dE/dx$ determination; as it does not suffer from inefficiencies of high multiplicity events seen in the collision experimental data.

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{cocktail.png}
\caption{Pad plane projection for a cocktail event in the TPC.}
\label{fig:cocktail}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{data.png}
\caption{Pad plane projection for a collision event in the TPC. Highlighted by red arrows are two regions of anode wires which had a reduced voltage of 1214 V. The voltage of the rest of the TPC anode wires are 1460 V. The reduction in voltage reduces the gain by a factor of about 10. }
\label{fig:data}
\end{figure}

\begin{figure*}[t]
\centering
\includegraphics[width=\linewidth]{dedx_compare}
\caption{The left panel shows the high gain stopping power vs low gain when the method of desaturation was not applied. In the right panel the desaturation technique was applied to the high gain region. The low gain does not suffer from saturation and represents the true $dE/dx$ value.}
\label{fig:lowvshigh}
\end{figure*}

\begin{figure*}[t]
\includegraphics[width=\linewidth]{cocktail_combine.png}
\caption{Uncorrected (left panel) and desaturated (right panel) cocktail data.}
\label{fig:cocktail_combine}
\end{figure*}

@@

Tracks which saturate pads in the high gain region are not saturated in the low gain region. By comparing the dE/dx values of these two sections, we can directly measure the success of the desaturation in the high gain regions using the method described above.

In Fig.~\ref{fig:lowvshigh}, the effect of saturation can be seen in the high gain region for the uncorrected data. For signals below 400 ADC/mm \footnote{Un-calibrated ADC channels in arbitrary units.} the electronics are not saturated, and therefore the high and low gain sections agree. The data starts to saturate above 400 ADC/mm in the high gain channels eventually reaching a plateau while the low gain sections are not saturated and provide the true $dE/dx$ values.
 After applying the desaturation method, the correlation between the high and low gain sections is restored, as seen in Fig.~\ref{fig:lowvshigh}. From this comparison, we infer that the correction works up to signals of 2000 ADC/mm, increasing the dynamic range by a factor of at least 5.

Comparing the low to high gain sections directly validates the desaturation technique, but the goal  of this exercise is to improve the particle identification (PID). In the following PID plots the red lines represent the most probable energy loss as given by Geant4 straggling functions. A linear calibration was performed to convert keV in Geant4 to ADC in the experiment given by $ADC/mm = 19\;keV/cm$.

There are pronounced PID lines of several particle species in both the uncorrected and corrected cocktail beam PID shown in the subplots of Fig.~\ref{fig:cocktail_combine}. Three ovals around a momentum of 1700~MeV/$c$/q and two near 900~MeV/$c$/q correspond to the three $B\rho$ settings injected into the TPC. The tails of the PID lines are resulting from the particles passing through the walls and other materials outside the main detector volume, therefore lowering the initial momentum.

The uncorrected data in Fig.~\ref{fig:cocktail_combine} shows the effects of saturation; the PID lines deviate from their theoretical expectations starting at around 400~ADC/mm eventually reaching a plateau. After applying the desaturation technique, we see a large improvement, most notably for the He and Li particles, which suffer the most from saturation. A more subtle improvement of the lighter particles, (p, d, t), can also be seen in the PID lines at lower momenta.

>>>>

\subsection{Gating grid noise subtraction}
Opening the gating grid essentially short adjacent wires together allowing them to reach equilibrium as fast as possible. In practice the impedance of both sides was not entirely matched properly and this caused an oscillating current to bounce back and forth in an under-dampened manner. This caused residual induced signals early in the time bucket spectrum which created and extra source of noise in the data. The ground grid shielded some of the noise, but the path to ground was on two ends and was not sufficiency for this frequency. Figure~\ref{fig:ggNoiseSub} shows in the upper panel the gating grid noise ADC time bucket spectrum for 2000 events in a given pad. The signal is stable within a pad and the mean value --shown as the red line-- can be calculated after averaging over several thousands events. The raw gating grid noise lasts for 100 tbs extending into the real data with a decreasing amplitude.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{ggNoise.png}
\caption{Gating grid noise profile and subtraction for a particular pad.}
\label{fig:ggNoiseSub}
\end{figure}

@@

Once the mean value response in each pad is calculated, we can use the gating grid noise profile runs to correct the real data and remove the noise of the gating grid. To show the best case scenario of this technique, the bottom panel of Fig.~\ref{fig:ggNoiseSub} shows the gating grid noise self correcting itself. We can see the pedestal is already subtracted naturally when using this technique, and the subtracted ADC spectra is centered around 0. Residual gating grid noise that was not canceled by the subtraction remains but is significantly reduced to a much smaller time bucket region. This happens in regions where the gating grid noise is changing rapidly, where the ADC value is dependent upon small jitters in the timing of the signal. All time buckets below 30 tb were cut away from the spectrum to get rid of the small region where the gating grid is still opening and to remove the residual noise.

%The average noise profile was constructed by averaging all the events in the gating grid runs, pad-by-pad; each pad had a slightly different noise profile since the impedance to each pad was slightly different. It is worth noting that the small electronics noise would be effectively canceled out by averaging over thousands of events. During the first stages of the software, the gating grid noise profile closest to that run would be subtracted from the raw data canceling the gating grid noise in that run.

@@

A cocktail of several light charged particles were produced (p,d,t,${}^{3}$He,${}^{4}$He), and measured in the TPC to provide a momentum calibration of the TPC. The magnetic and slit settings of the dipoles in the BIGRIPS spectrometer was set so that the measured momentum resolution of the beam was $\delta p/p$ < 1\%. Two magnetic rigidity settings were studied, with an empty target. A thick Aluminum target was used to provide a slightly lower energy point for part of the lower rigidity setting, effectively creating three calibration points over several particle species. The production of certain particles (t,${}^{3}$He), produced too few counts to make a good measurement, and in the high momentum rigidity setting protons could not propagate down the line and there were no counts.

Since the expected momentum resolution resulting from the spectrometer was less than 1\%, the observed momentum resolution measured by the TPC is a good representation of the combined momentum resolution of the software and TPC (intrinsic detection) system. The momentum resolution depends on several factors such as the particle's angle, momentum, charge, track multiplicity, etc. This calibration beam represents and ideal situation where the track was parallel to the pad plane and only one particle was measured at one time. The energy loss resolution can also be directly measured since each energy setting represents a monochromatic source of each particle species, which has a well defined energy loss distribution. An average momentum resolution of $dp/p = 2\%$ and the energy loss resolution of $d\langle dE/dx \rangle/\langle dE/dx \rangle = 5\%$ was measured for particles ranging from $H$ to ${}^{4}$He, over the range of momenta measured in the calibration beam as summarized in Table~\ref{tb:momresolution}.

Since the magnetic dipole setting of the BIGRIPS spectrometer defined the energy of each particle type, we can calculate the expected momenta of each particle species measured. Small corrections were propagated using LISE++ software which can calculate the energy loss through several materials in the beam line. These corrections resulted in a small change in the momenta. The measured momenta of the calibration beam differed significantly from the expected values as seen in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}. This effect is attributed to inhomogenatities in the magnetic field which introduces electron drift velocity in the direction of $\vec{E}\times\vec{B}$ direction as seen from Eq.\ref{eq:elecdrift}. The $\vec{E}\times\vec{B}$ drift velocity causes the electron trajectories to shift toward the +x-axis in the TPC coordinates causing particles of positive charge (going in the -x-axis) to have a higher measured momenta than in reality. The disagreement in measured and expected momenta is upwards of ~5\% difference in the higher momentum calibration settings. The details of the correcting for the $\vec{E}\times\vec{B}$ effect will be discussed in the later Section~\ref{sec:spacecharge} in a more general way which also includes correction for the space charge, which this cocktail beam does not have. The same correction technique was applied here in the special case of zero space charge, which is the special case of only having $\vec{E}\times\vec{B}$ components.

The values under the corrected column seen in Tables~\cref{tb:cocktail100tar,tb:cocktail100,tb:cocktail300}, represent the data corrected for the $\vec{E}\times\vec{B}$ effect. A significant improvement is seen in the high momentum setting going from around ~5\% disagreement to within ~1\% agreement in the corrected data. For the lower momentum settings (Tables~\cref{tb:cocktail100tar,tb:cocktail100}), protons see a slight improvement of about ~1\% where as the deuterons are over corrected in both settings. The level of agreement of the all corrected values is still within the estimated momentum resolution of the TPC.

@@

%Apendix of the beam line elements
%apendix of the settings of the bigrips elments
%Picture of cocktail before and after ExB effect
%Table of LISE++ expected cocktail energies ridigity setting of dipole magnets (reference big rips line)

\subsection{Electronics calibration}
The ADC channel number of the electronics was calibrated by measuring the response of each channel to an input pulse supplied by a pulse generator. The pulse was distributed to all the electronics channels by pulsing the ground plane for a range of input voltages. This distributed the pulse evenly across the entire pad plane. The input voltage is plotted as a function of the measured ADC channel in Fig.~\ref{fig:gaincalib} for every channel. The small variation in each channel can be seen as the wide band around each measurement point. A linear fit is performed to get the best fit line which provides a reference line which each channel is calibrated to. The right panel shows the resulting distribution of channels after calibration. This is a relative calibration technique with the intent to calibrate the varying gains in each channel relative to one another.

@@

As mentioned earlier, the anode wires were separated into 14 independently biased sections. The high voltage of sections 12 and 14 were reduced during the experiment due to high currents being observed on the wires. The wire section voltages were lowered once and adjusted once again. Out of all the runs used in the analysis in this thesis \ref{tb:runList}, the anode sections 12 and 14 were lowered to \SI{1085}{\volt} for runs 2272-2371 and set to \SI{1214}{\volt} for all the other runs. By lowering the voltage on these anode wires, the gas gain is lowered as compared with all the other anode wire plane sections which operate at \SI{1460}{\volt}. To account for the drop in gain, in the software we increase the gain of the pads which lie above these anode wires to match the gain in the other channels. To do so, we compare the energy loss in the high gain sections to the low gain sections to get the relative calibration factor.

@@

Figure~\ref{fig:highlowcal} shows the correlation plot of $\langle dE/dx\angle$ of the high versus the low gain sections. The effects of the high gain channel saturation can around a value of 400 ADC\si{\per\milli\metre} and plateaus, where as the low gain does not. A linear fit was performed for the calibration  between the high gain channel $h_c = G\cdot l_c$, where $l_c$ is the value of the low gain channel and $G$ is the gain factor. In the case where the low anode voltage was \SI{1214}{\volt}, the gain factor was 9.8 where as in the case of \SI{1085}{\volt}, the gain factor was ???.  The calibration factor was applied inside of the decoder task. A map of the channels which are impacted by the low anode wire voltages and the relevant gain was input into the software. In the decoder task the raw ADC spectra was multiplied by the input gain for a given channel. The thresholds of those channels also were multiplied by the same factor since the noise levels were also artificially magnified by the same gain factor.

\subsection{Extending the dynamic range of the Electronics}
Using a TPC for measurements of HIC in nuclear physics presents a different set of challenges as opposed to higher energy experiments. Typically in higher energy experiments the charge of particles is Z$e$, where Z=1. Also, the particles are traveling at higher energies in which the energy losses are near or close to the minimum in the energy loss curve. The dynamic range of electronics in such experiments can cover a wide range of particle energies in the energy loss curve. In nuclear HICs, we are interested in measured particles ranging from Z=1-3 resulting from the collision, and even higher in some applications. As seen in Eq.~\ref{eq:bb}, the energy loss is proportional to $Z^2$; the energy values in nuclear HIC of intermediate beam energies (around \SI{300}{\MeVA}) in which the resulting particles are at much lower energies which are in the $1/\beta^2$ region. The energy range and particle types covered by the electronics are significantly limited by the dynamic range limitations of the electronics as the charge of a particle increases and the velocity decreases.

Several TPCs have tried to address this issue by having regions of low and high gain, either in amplification gain or in electronics gain. Here we run into the issue that high velocity particles will have little to no signal in the low gain regions. At lower velocities, particles will deposit much charge over the low gain regions but will saturate the high gain regions, whose charge values are usually lost. Only within the dynamic range will tracks have the best momentum resolution, outside the dynamic range clusters will be missing in the sections which depend on the track velocity, lowering the momentum resolution of tracking. There are ongoing efforts in the nuclear community to develop new electronics which hope to mitigate these persistent issues in TPC electronics CITE HERE, by being able to switch to a lower gain value when the maximum range is reached. Though, it is quite useful to develop a software technique which may extend the dynamic range of TPC electronics without the use of external hardware, especially in experiments which have already been performed with preexisting electronics technologies.

In TPCs, the effective dynamic range can be very different from the single channel dynamic range depending on the application. Typically TPCs are operated inside of a magnetic field for reconstructing momentum of each particle, which requires sub-millimeter precision in the position determination of the track path. This is achieved by averaging the charge distribution over several pads. Therefore the effective dynamic range is related to the relative charge values of adjacent pads (between the center pad and outside pads). For example, to measure minimum ionizing particles the signal to noise ratio of the pad with the smallest charge in the distribution should be some reasonable value above noise, say at least 6:1 signal-to-noise ratio (SNR). From Section \ref{sec:prf}, we know the central pad in the cluster holds 80\% of the total charge, where as the two adjacent pads each hold the remaining 10\%. If we require the adjacent pads to have a SNR of 6:1, then the central pad would have a SNR of 50:1. Considering this is the SNR for minimum ionizing particles, and the maximum SNR in a channel is 800:1, this means the effective dynamic range in the TPC is roughly 16 times that of minimum ionizing particles, if we intend to have adequate position and tracking resolution.

The dashed lines and vertical blue bar in Fig.~ \ref{fig:intro} are separated by a factor of 20, representing the typical effective dynamic range in a TPC. This dynamic range estimate should be regarded as approximate because the energy loss fluctuates significantly about the most probable energy loss, with a long ``Landau'' like tail, as described by Bichsel \cite{bichsel}. Nevertheless, the blue dashed lines and vertical blue bar illustrate that the range of energy losses sampled in a fixed gain readout system is limited. One can change the gain and shift the energy loss range that can be sampled, but the dynamic range itself cannot be increased.

@@

The rapid increase in the stopping powers at low momentum illustrates the degree to which the effective dynamic range can be exceeded and highlights the problems encountered in studies of intermediate HICs, in which light particles with low momenta are abundantly produced along with highly charged particles. Similar problems are encountered when TPCs are used as active targets in direct reaction studies with rare isotope beams \cite{pattpc}. Several techniques have been employed to increase the observable range of energy losses; typically by lowering the electronics gain of selected readout channels, or by changing the gas amplification at the readout plane in certain areas of the TPC \cref{eos,pattpc}. The results of changing the gas-gain, or the electronics gain, are rather similar in that reducing the gain to sample a range of higher energy loss makes the TPC effectively blind to minimum ionizing particles in  the regions of lower gain.

@@

The pad response function of the TPC was extracted from non-saturating hits and clusters in tracks at various track crossing angles. As in Fig.~\ref{fig:topview}, we postulate that the PRF is a function of the total charge deposited in a cluster $Q = \sum_i q_i$, and the difference in position of the center of the $i^{th}$ pad, $x_i$, to the mean position $\bar{x} = \sum_i x_i q_i/Q$, defined as $\lambda_i = x_i-\bar{x}$. The PRF is simply defined as the charge fraction of each pad as a function of $\lambda$, as shown in Equation \ref{eq:prf}.

@@

\begin{comment}
\begin{figure}[ht!]s
\begin{overpic}[width=\linewidth]{fig5.pdf}
\put(61,55){\contour{white}{ PRF${}_{\mathrm{Gaus}}(\lambda)$ eq. \ref{eq:gaus}  }}
\put(61,49){\contour{white}{ PRF${}_{\mathrm{Gatti}}(\lambda)$ eq. \ref{eq:gatti} }}
\end{overpic}
\caption{Experimental pad response function of many events for a crossing angle of $85^{\circ} < \theta \leq 90^{\circ}$.  }
\label{fig:expprf}
\end{figure}
\end{comment}

The shape of the PRF depends on the crossing angle of the track \cite{gatti}, which determines how wide the charge is distributed along the wire. Figure~\ref{fig:prfpimData} shows the PRF of $\pi^-$ tracks versus the crossing angle $\theta$ of the track. The PRF gets wider starting from $90^{\circ}$  and until where we switch clustering directions at $\ang{45}$; if we did not switch clustering directions the PRF would become wider until it was a uniform distribution and there was no position resolution. Switching from $x$ to the $z$ direction clustering the opposite trend is seen where the PRF becomes narrower  going from $45^{\circ}$ to $0^{\circ}$, as the position resolution gets better.

@@

A Gaussian fit was performed were performed to the experimental data with  $\ang{5}$ width bins from $\ang{0} < \theta \leq \ang{90}$. Figure~\ref{fig:normsigma} shows the two parameters resulting from fitting the the Gaussian function given in Eq.~\ref{eq:prfgaus} are plotted versus $\theta$. A $4^{th}$ order polynomial fit of these parameters allowed for interpolating for any given $\theta$ value as seen by the black line.

\subsection{Method of Desaturation}

@@

We will use the term ``desaturation'' to describe the process of correcting the charge values of the saturated pads. Figure~\ref{fig:satpad} shows a typical situation of saturated signals in a given cluster of hits. When an avalanche causes a large enough induced signal, it is the pads directly underneath which collect the largest charge becoming saturated; denoted here as $q_{2\textprime}$ and $q_{3\textprime}$. Pads further away collect less charge and typically are not saturated; the non-saturated pads are detonated here as $q_{1}$ and $q_{4}$. Though the true charge value in the saturated channels are lost, we know that the distribution of all charges must follow the PRF which is a fixed feature of all TPCs, and has been experimentally measured above. From the tracking information, we know a given clusters crossing angle and can interpolate to get the corresponding parameters for the PRF described in Fig.~\ref{fig:normsigma}.

We assume the distance of each pad to the track, $\lambda_i$, is fixed, defining the fraction of charge each pad receives as given by the $PRF(\lambda_i)$ function. To determine the best estimate for the charge values of each saturated pad, a chi squared function is minimized,

\begin{equation}\label{eq:chi}
\chi^2 = \sum_i \frac{(q_i^{\mathrm{obs}} - q_i^{\mathrm{expect}})^2}{q_i^{\mathrm{expect}}},
\end{equation}

where $q_i^{\mathrm{obs}}$ are the non-saturated charges and $q_i^{\mathrm{expect}}$ are the charge values expected from the PRF calculated as $q_i^{\mathrm{expected}} = Q\cdot PRF(\lambda_i)$. The saturated charge values $q_{i}^{\textprime}$ are treated as unknown variables and are allowed to vary in the $\chi^2$ minimization; the values calculated by the minimizer are added to the total expected charge $Q$.

@@

Tracks which saturate pads in the high anode wire voltage region are not saturated in the low anode voltage region. By comparing the $\langle dE/dx\rangle$ values of these two sections, we can directly measure the success of the desaturation in the high gain regions. While this desaturation technique avoids the need to lower the gain of any region, the low anode voltage region proved to be a direct measurement of the success of this technique.

Figure~\ref{fig:lowvshigh} shows the the $\langle dE/dx\rangle$ values of the high gain region compared with the calibrated low gain region. The effect of saturation can be seen in the high gain region for the uncorrected data above values of 400 ADC/mm where the values plateau, where as the low gain region still returns accurate values\footnote{Un-calibrated ADC channels in arbitrary units.} Below this value the electronics are not saturated, and therefore the high and low gain sections agree.

 After applying the desaturation method, the correlation between the high and low gain sections is restored, as seen in Fig.~\ref{fig:lowvshigh}. From this comparison, we can approximate that the correction has corrected the high gain sections to agree with low gain sections to values of 2000 ADC/mm, increasing the dynamic range by a factor of at least 5.

The success of the desaturation becomes more clear when looking at the particle identification (PID) lines. In the following PID plots the red lines represent the most probable energy loss as given by Geant4 straggling functions. A linear calibration was performed to convert keV in Geant4 to ADC in the experiment given by $ADC/mm = 19;keV/cm$.

There are pronounced PID lines of several particle species in both the uncorrected and corrected cocktail beam PID shown in the subplots of Fig.~\ref{fig:cocktail_combine}. The uncorrected data in the left panel shows the effects of saturation; the PID lines deviate from their theoretical expectations starting at around 400~ADC/mm eventually reaching a plateau. After applying the desaturation technique -- in the right panel-- we see a large improvement, most notably for the He and Li particles, which suffer the most from saturation and even ${}^{6}Li$ and ${}^{7}Li$ can be separated. A more subtle improvement of the lighter particles, (p, d, t), can also be seen in the PID lines at lower momenta.

A similar effect is seen in the experimental collision data. In Figure~\ref{fig:data_combine} though the PID suffers from more background and inefficiencies than the cocktail beam we see a similar improvement in the PID lines when comparing before -- the left panel-- and after applying desaturation -- the right panel. Notably the largest improvement is the separation of particle species at lower momenta and the separation of the Li species into ${}^{6}$Li and ${}^{7}$Li. In these regions, there was little to no PID resolution before desaturation.

The dynamic range was extended by at least a factor of 5, as demonstrated by the improved PID lines, and quantified by direct comparison to low gain sections of the TPC.

%Give some failures of the assumptions made
